```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = F,
                      comment = F, 
                      warning = F, 
                      message = F)
```


```{r}
require(umap)
require(ggplot2)
require(gridExtra)
```


# UMAP

UMAP is an algorithm for non-linear dimension reduction. UMAP learns the 
manifold structure of the input data, and finds a **low dimensional embedding** 
that preserves the topological structure of that manifold.

Here we will simulate data and generate 2D UMAP


## Data simulation

  * $N$=100 cells
  * $F$=2 features (F1 and F2)
  * Gene expression matrix $Y^{N \times F}$
      
      * $Y^{[1:50, F1]} \sim N(0, 1)$
      * $Y^{[51:100, F1]} \sim N(5, 2)$
      * $Y^{[1:100, F2]} \sim N(0, 1)$


### F1 and F2 distribution


```{r sim100}
# simulate 100 expression in cells with 2 features
# feature 1: first 50 cell expressions ~ N(0, 1)
# feature 1: second 50 cell expressions ~ N(5, 2)
# feature 2: all 100 cell expressions ~ N(0, 1)
set.seed(seed = 7612)
y <- cbind(c(rnorm(n = 50, mean = 0, sd = 1), 
             rnorm(n = 50, mean = 5, sd = 2)),
           rnorm(n = 100, mean = 0, sd = 1))
rownames(y) <- paste0("C-", 1:nrow(y))
colnames(y) <- paste0("F-", 1:ncol(y))
```



```{r fig_data, fig.width=8, fig.height=4}
gridExtra::grid.arrange(
  nrow=1,
  ggplot(data = data.frame(y))+
  geom_point(aes(x = 1:nrow(y), y = y[, 1]))+
  theme_bw()+
  ylab(label = "F1")+
  xlab(label = "Cells 1 to 100"),
  ggplot(data = data.frame(y))+
  geom_point(aes(x = 1:nrow(y), y = y[, 2]))+
  theme_bw()+
  ylab(label = "F2")+
  xlab(label = "Cells 1 to 100"))
```




```{r umap-orig}
# build umap
u <- umap(d = y)

# data.frame of UMAP coordinates
m <- data.frame(u$layout)
m <- cbind(m, data.frame(u$data))
m$G <- rep(x = c("1-50", "51-100"), each = 50)
```

```{r fig1, fig.width=9, fig.height=4}
gridExtra::grid.arrange(
  nrow=1,
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = G))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top"),
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = F.1))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    scale_color_distiller(palette = "Spectral"),
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = F.2))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    scale_color_distiller(palette = "Spectral"))
```


## UMAP parameters

UMAP has a lot of parameters. See full list below.

```{r}
u.conf <- umap::umap.defaults
print(unlist(u.conf))
```

### *n_neighbors*  (default n_neighbors: 15)
*n_neighbors* controls the resolution at which UMAP learns the manifold 
structure of the data.

If *n_neighbors* is set to a too low value, UMAP will zoom into local 
neighborhood when attempting to learn the manifold structure of the data. 
The opposite will happen if *n_neighbors* is set to a too high value.


Lets set *n_neighbors*=3 and look at the resulting UMAP. As expected, we see 
that UMAP focuses on learning from local structure and misses the global 
structure. As a result we get stringy UMAP layout.

```{r orig_umap, fig.width=8, fig.height=4}
# change config of umap
u.conf <- umap::umap.defaults
u.conf$n_neighbors <- 3

u.temp <- umap(d = y, config = u.conf)

# data.frame of UMAP coordinates
m <- data.frame(u.temp$layout)
m <- cbind(m, data.frame(u$data))
m$G <- rep(x = c("1-50", "51-100"), each = 50)


gridExtra::grid.arrange(
  nrow=1,
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = G))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top"),
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = F.1))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    scale_color_distiller(palette = "Spectral"),
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = F.2))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    scale_color_distiller(palette = "Spectral"))
```

The opposite should happen for e.g. *n_neighbors*=100, the overal (global) 
structure should be captured better but at the loss of some of the finer local structure compared to the initial example where we set *n_neighbors*=15. As 
there are only 100 cells in our data, *n_neighbors*=15 was already high enough 
and we do not see that much of difference in the newly built UMAP.

```{r n_neighbors, fig.width=8, fig.height=4}
# change config of umap
u.conf <- umap::umap.defaults
u.conf$n_neighbors <- 100

u.temp <- umap(d = y, config = u.conf)

# data.frame of UMAP coordinates
m <- data.frame(u.temp$layout)
m <- cbind(m, data.frame(u.temp$data))
m$G <- rep(x = c("1-50", "51-100"), each = 50)

gridExtra::grid.arrange(
  nrow=1,
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = G))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top"),
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = F.1))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    scale_color_distiller(palette = "Spectral"),
  ggplot(data = m)+
    geom_point(aes(x = X1, y = X2, col = F.2))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    scale_color_distiller(palette = "Spectral"))
```

### *min_dist* (default min_dist: 0.1)
*min_dist* determines the compactness/clumpiness of the data points as it 
provides the minimum possible distance that may exist between any two points.

Below two panels show UMAPs with *min_dist*=$10^{-5}$ (left panel) and 
*min_dist*=0.99 (right panel)

```{r min_dist, fig.width=8, fig.height=4}
set.seed(seed = 7612)

# change config of umap
u.conf <- umap::umap.defaults
u.conf$min_dist <- 10^(-5)
u.temp <- umap(d = y, config = u.conf)
# data.frame of UMAP coordinates
m1 <- data.frame(u.temp$layout)
m1 <- cbind(m1, data.frame(u.temp$data))
m1$G <- rep(x = c("1-50", "51-100"), each = 50)



# change config of umap
u.conf <- umap::umap.defaults
u.conf$min_dist <- 0.99
u.temp <- umap(d = y, config = u.conf)
# data.frame of UMAP coordinates
m2 <- data.frame(u.temp$layout)
m2 <- cbind(m2, data.frame(u.temp$data))
m2$G <- rep(x = c("1-50", "51-100"), each = 50)


gridExtra::grid.arrange(
  nrow=1,
  ggplot(data = m1)+
    geom_point(aes(x = X1, y = X2, col = G))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    ggtitle(label = "min_dist=10^(-5)"),
  ggplot(data = m2)+
    geom_point(aes(x = X1, y = X2, col = G))+
    xlab(label = "UMAP_1")+
    ylab(label = "UMAP_2")+
    theme_bw()+
    theme(legend.position = "top")+
    ggtitle(label = "min_dist=0.99"))
```

### *n_components* (default n_components: 2)
As in other dimension reduction algorithms, UMAP can perform dimension 
reduction to a specific number of dimensions. This is regulated by the 
parameter *n_components*.


```{r n_components, fig.width=4, fig.height=4}
set.seed(seed = 7612)

# change config of umap
u.conf <- umap::umap.defaults
u.conf$n_components <- 1
u.temp <- umap(d = y, config = u.conf)
# data.frame of UMAP coordinates
m <- data.frame(u.temp$layout)
m <- cbind(m, data.frame(u.temp$data))
m$G <- rep(x = c("1-50", "51-100"), each = 50)
m$i <- 1:nrow(m)

ggplot(data = m)+
    geom_point(aes(x = i, y = u.temp.layout, col = G))+
    xlab(label = "cells 1 to 100")+
    ylab(label = "UMAP_1")+
    theme_bw()+
    theme(legend.position = "top")+
    ggtitle(label = "n_components=1")
```


### *metric* (default metric='Euclidean')

This parameters determines how distances are computed between the cells. Many
different metrics are available



